Assessing Infant Mortality Rate: Problems stemming from Household Living Conditions, Women's Education and Health

### 2022-04-27 outputs/paper/paper.Rmd
update to paper changing grammar in intro

---
title: "Assessing Infant Mortality Rate: Problems stemming from Household Living Conditions, Women's Education and Health"
subtitle: "Trends in India from the 1998 Demographic Health Survey"
author: 
  - Jacob Yoke Hong Si
thanks: "Code and data are available at: https://github.com/jacobyokehongsi/Assessing_Infant_Mortality_Rate_in_1998_India"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "What areas can be improved in order to promote the well-being of women in India and hence, reduce the infant mortality rate? Utilizing the data from the 1998-1999 India National Family Health Survey provided by the Demographic and Health Survey (DHS) program, we look to depict the demographics of Indian women and infants in different states of India. We have found that the root cause of poor infant mortality rates stems from having poor living conditions that affect the likelihood of women to attain education and understand the importance of antenatal care and birth delivery assistance. We also explore other factors such as inheritable traits (unhealthy body weight and anaemia disease) as well as an infant's diet. These factors are crucial in the development of an infant and the reduction of the infant mortality rate. \\par \\textbf{Keywords:} demographics, health, living conditions, infant mortality rate, education, antenatal care, body weight, anaemia, diet."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

@@

The National Family Health Survey (NFHS-2: DHS program in India) on the state findings in India assesses key demographics of women and infants including their education, healthcare and living conditions. These factors are of utmost importance in the growth of a human being. Women that are highly educated tends to pay more attention to their own health leading to a happier and healthier lifestyle. Furthermore, these characteristics of a women would be passed down generations, affecting the health conditions of their infant.

@@

The data was obtained from India’s second National Family Health Survey (NFHS-2) on women's and infant's health. The survey was conducted in 1998-1999 by the International Institute for Population Sciences (IIPS) using the NFHS questionnaire where households would be interviewed. The questionnaire was devised during workshops held in Mumbai where experts in population and health fields would partake in creating the questions. The target population was a sample of more than 90,000 ever-married women age 15–49 as well as infant born in the three years preceding the survey. The list of villages is used as a sampling frame and first stratified through several variables. Next, the participants were selected using systematic sampling with equal probability from a household list in each village area. The overall individual response rate was 96% with a total of 89,199 eligible women that are part of the 91,196 interviewed households.

@@

## Survey Frame and Sampling Method
As mentioned in the previous section, participants of this survey were chosen via stratified sampling and systematic sampling. The survey frame that was created is as follows. The list of villages were first stratified by geographic location, then continuous regions and lastly further stratified using selected variables such as village size, female literacy and so on. Each state in India was stratified individually using these variables with a limit of 6 stratas for small states, 12 stratas for medium size states and 15 stratas for large states. From the list of stratified villages, they are then sampled systematically with probability proportional to the 1991 Census population of the village. The villages that have fewer than five households are not included in the sampling frame. Lastly, the individuals to be interviewed were selected with equal probability from the household list in each village area using systematic sampling.

## Key features
The raw data of the state findings includes 70 variables which were obtained from the questionnaire responses. Key features part of the raw data include education standards, accessibility to electricity, drinking water and toilets, infant mortality rate, mothers receiving antenatal care, anaemia statuses (mothers and infant), weight statuses (mothers and infant) and the infant's diet.

We used R [@citeR] and R packages tidyverse [@tidy] and pdftools [@pdftools] to extract, preprocess and analyze the data. When preprocessing the data, we also used R package pointblank [@pointblank] to put together tests for class and content to ensure that the dataset passes. The key features analyzed in this paper is shown in table 1 and 2, which was made using R package knitr [@knitr].

@@

## Strengths and Weaknesses
### Strengths
The primary strengths of this survey is the myriad of variables describing participants and households, as well as the large sample size where more than 99 percent of India's female population is surveyed. Additionally, all Indian states except for one is surveyed, covering the majority of India. The field data collected by interviewing teams were inputted into microcomputers to produce field-check tables, ensuring that errors are prevented when eliciting information and filling out questionnaires. The large amount of variables allows for thorough data exploration of the relationships between them while the ample amount of data allows concrete conclusions to be drawn. Furthermore, the sampling methodologies is explained in high detail as well as a comprehensive fact sheet put together to display key data from the survey.

### Weaknesses
One of the drawbacks of the survey is that survey findings were obtained in all states but one, that is Tripura, due to the delay in fieldwork based on a local problem. Additionally, data is unable to be obtained from areas with union territories. Nonsampling errors are also present where surveyors were unable to locate and interview the correct household and the misunderstanding of the questionnaire by the interviewer or respondent. In addition, the collection of anaemia data was hindered since some female participants refused to partake in the anaemia testing and/or have their infant tested as well. These aforementioned weaknesses results in missing data thus, the inference we will make could be less diverse.

# Results
The figures were made using R [@citeR], R package tidyverse [@tidy] and Lucidchart [@lucidchart]. The variables explored provides numeric values that varies between the states of India.

@@

Figure \@ref(fig:dag) represents a directed acyclic graph that will outline the relationships between the variables explored in the paper.

@@

## Mothers that receive antenatal check-up, antenatal care and birth delivery methods
Here, we explore how mothers that have antenatal check-up affect the antenatal care (supplements and injections) they receive for births three years preceding the survey.

@@

### Birth Delivery Methods
```{r antemed, echo=FALSE, warning=FALSE, message=FALSE, fig.align="center", fig.cap="Relationship between mothers receiving at least one antenatal check-up and deliveries in medical institutions (in percent)"}
data %>% 
  ggplot(aes(x = antenatal, y = deli_med_inst)) +
  geom_point() +
  theme_minimal() +
  labs(x = "Mothers receiving at least one antenatal check-up",
       y = "Deliveries in medical institutions") + 
  geom_smooth(method='lm')
```

@@

# Discussion
The ultimate goal of the paper is to assist policy makers in implementing appropriate and efficient socio-economic policies to help promote the steady growth of infants and reduce infant mortality rate in India. We will first discuss the root of the issue which is the household living environment, women's education and their healthcare decisions which lead them to attend antenatal check-ups. Then, we examine the numerous factors that are affected by conducting antenatal check-ups which include antenatal supplements and injections, as well as birth delivery methods. These are some of the medical factors that directly impact the infant mortality rate. Additionally, we also look at the importance of infant's meals for their growth. Lastly, we examine inheritable traits such as anaemia and unhealthy body weight that could be passed on from mother to infant and how they impact the infant mortality rate. The full data can be found in Appendix B.

>>>>

---
title: "Assessing Infant Mortality Rate: Problems stemming from Household Living Conditions, Women's Education and Health"
subtitle: "Trends in India from the 1998 Demographic Health Survey"
author: 
  - Jacob Yoke Hong Si
thanks: "Code and data are available at: https://github.com/jacobyokehongsi/Assessing_Infant_Mortality_Rate_in_1998_India"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "What areas can be improved in order to promote the well-being of women in India and hence, reduce the infant mortality rate? Utilizing the data from the 1998-1999 India National Family Health Survey provided by the Demographic and Health Survey (DHS) program, we look to depict the demographics of Indian women and infants in different states of India. We have found that the root causes of poor infant mortality rates stem from having poor living conditions that affect the likelihood of women to attain education and understand the importance of antenatal care and birth delivery assistance. We also explore other factors such as inheritable traits (unhealthy body weight and anaemia disease) as well as an infant's diet. These factors are crucial in the development of an infant and the reduction of the infant mortality rate. \\par \\textbf{Keywords:} demographics, health, living conditions, infant mortality rate, education, antenatal care, body weight, anaemia, diet, birth delivery assistance."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

@@

The National Family Health Survey (NFHS-2: DHS program in India) on the state findings in India assesses key demographics of women and infants including their education, healthcare and living conditions. These factors are of utmost importance in the growth of a human being. Women that are highly educated tend to pay more attention to their own health leading to a happier and healthier lifestyle. Furthermore, these characteristics of a women would be passed down generations, affecting the health conditions of their infants.

@@

The data was obtained from India’s second National Family Health Survey (NFHS-2) on women's and infants' health. The survey was conducted in 1998-1999 by the International Institute for Population Sciences (IIPS) using the NFHS questionnaire where households would be interviewed. The questionnaire was devised during workshops held in Mumbai where experts in population and health fields would partake in creating the questions. The target population was a sample of more than 90,000 ever-married women age 15–49 as well as infants born in the three years preceding the survey. The list of villages are used as a sampling frame and first stratified through several variables. Next, the participants were selected using systematic sampling with equal probability from a household list in each village area. The overall individual response rate was 96% with a total of 89,199 eligible women that were part of the 91,196 interviewed households.

@@

## Survey Frame and Sampling Method
As mentioned in the previous section, participants of this survey were chosen via stratified sampling and systematic sampling. The survey frame that was created is as follows. The list of villages were first stratified by geographic location, then continuous regions and lastly further stratified using selected variables such as village size, female literacy and so on. Each state in India was stratified individually using these variables with a limit of 6 stratas for small states, 12 stratas for medium size states and 15 stratas for large states. From the list of stratified villages, they were then sampled systematically with probability proportional to the 1991 Census population of the village. The villages that had fewer than five households were not included in the sampling frame. Lastly, the individuals to be interviewed were selected with equal probability from the household list in each village area using systematic sampling.

## Key features
The raw data of the state findings included 70 variables which were obtained from the questionnaire responses. Key features part of the raw data included education standards, accessibility to electricity, drinking water and toilets, infant mortality rate, mothers receiving antenatal care, anaemia statuses (mothers and infants), weight statuses (mothers and infants) and the infants' diet.

I use R [@citeR] and R packages tidyverse [@tidy] and pdftools [@pdftools] to extract, preprocess and analyze the data. When preprocessing the data, I also use R package pointblank [@pointblank] to put together tests for class and content to ensure that the dataset passes. The key features analyzed in this paper are shown in table 1 and 2, which is made using R package knitr [@knitr].

@@

## Strengths and Weaknesses
### Strengths
The primary strengths of this survey was the myriad of variables describing participants and households, as well as the large sample size where more than 99 percent of India's female population was surveyed. Additionally, all Indian states except for one was surveyed, covering the majority of India. The field data collected by interviewing teams were inputted into microcomputers to produce field-check tables, ensuring that errors were prevented when eliciting information and filling out questionnaires. The large amount of variables allowed for thorough data exploration of the relationships between them while the ample amount of data allowed concrete conclusions to be drawn. Furthermore, the sampling methodologies was explained in high detail as well as a comprehensive fact sheet put together to display key data from the survey.

### Weaknesses
One of the drawbacks of the survey was that survey findings were obtained in all states but one, that wa Tripura, due to the delay in fieldwork based on a local problem. Additionally, data was unable to be obtained from areas with union territories. Nonsampling errors were also present where surveyors were unable to locate and interview the correct household and the misunderstanding of the questionnaire by the interviewer or respondent. In addition, the collection of anaemia data was hindered since some female participants refused to partake in the anaemia testing and/or have their infants tested as well. These aforementioned weaknesses results in missing data thus, the inference we will make could be less diverse.

# Results
The figures are made using R [@citeR], R package tidyverse [@tidy] and Lucidchart [@lucidchart]. The variables explored provides numeric values that varies between the states of India.

@@

Figure \@ref(fig:dag) represents a directed acyclic graph that will outline the interrelationships between variables affecting Infant Mortality Rate.

@@

## Mothers that receive antenatal check-up, antenatal care and birth delivery assistance
Here, we explore how mothers that have antenatal check-up affect the antenatal care (supplements and injections) they receive for births three years preceding the survey.

@@

### Birth Delivery Assistance
```{r antemed, echo=FALSE, warning=FALSE, message=FALSE, fig.align="center", fig.cap="Relationship between mothers receiving at least one antenatal check-up and deliveries in medical institutions (in percent)"}
data %>% 
  ggplot(aes(x = antenatal, y = deli_med_inst)) +
  geom_point() +
  theme_minimal() +
  labs(x = "Mothers receiving at least one antenatal check-up",
       y = "Deliveries in medical institutions") + 
  geom_smooth(method='lm')
```

@@

# Discussion
The ultimate goal of the paper is to assist policy makers in implementing appropriate and efficient socio-economic policies to help promote the steady growth of infants and reduce infant mortality rate in India. We will first discuss the root of the issue which is the household living environment, women's education and their healthcare decisions which lead them to attend antenatal check-ups. Then, we examine the numerous factors that are affected by conducting antenatal check-ups which include antenatal supplements and injections, as well as birth delivery assistance. These are some of the medical factors that directly impact the infant mortality rate. Additionally, we also look at the importance of infant's meals for their growth. Lastly, we examine inheritable traits such as anaemia and unhealthy body weight that could be passed on from mother to infant and how they impact the infant mortality rate. The full data can be found in Appendix B.
